State & Transition

### 2023-04-25 methods.Rmd
Final edits for submission. Edits for resubmission to include double spacing and line numbers

**TODO Dongchen: consider adding a conceptual diagram to explain the framework and data**

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use both to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) and sometimes model parameters. These updated variables are then used as the initial conditions for the next Forecast round.

Within the EnKF, the probabilistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error. The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model. In EnKF the forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$. In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200. Therefore $\mu_{f,t}$ is a vector of length 3 and $P_{f,t}$ is a $3 \times 3$ covariance matrix. If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space. An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix comes from (1) the mechanistic understanding embedded in our process model and (2) the spatial covariance in the forecast inputs (initial conditions, parameters, drivers, process error), rather than an empirical semivariogram.

The Analysis step in the EnKF is the same as that in the standard Kalman Filter. Because the forecast step from t-1 to t occurs prior to the observed data at time t, $Y_t$, the forecast of our three state variables is used as an informative _prior_ distribution and is assumed to follow a multivariate Normal (MVN) distribution with mean $\mu_{f,t}$ and covariance $P_{f,t}$. The EnKF also acknowledges that the observed data at time t, $Y_t$, is measured imperfectly. Observation error is assumed to follow a MVN distribution with an observation error covariance matrix $R$. Furthermore, it is assumed that $R$ is known, which is not unreasonable as many derived data products and sensors report uncertainties, and uncertainties from field data can often be inferred from sampling theory or Monte Carlo resampling / bootstrapping. In many cases new data do not exist for every state variable (e.g. we may observe leaves and wood but not soil C) and so an observation operator matrix, $H$ is also introduced, which in most cases is just a $n \times m$ indicator matrix matching the observed $Y$'s (n=2 in this example) with the vector of state variables (m=3 in this example). Indeed, one of the powerful features of data assimilation is its ability to update unobserved state variables based on the covariances between observed and unobserved variables. Putting this all together, Bayes' Theorem can then be used to update our MVN forecast prior, given a MVN Likelihood of $Y_t$ with covariance $R$, giving an analytically-tractable conjugate MVN posterior with mean $\mu_{a,t}$ and covariance $P_{a,t}$:

@@

The subscript $a$ is used to refer to the Analysis posterior and distinguish this from the prior forecast $f$ mean and covariance for the same timestep. While the matrix notation in these equations can make these solutions look complicated, the underlying meaning is accessible. First, the precision of the analysis (i.e. 1/variance) is the sum of the observation precision and the forecast precision, meaning that the analysis is always more precise than the data or forecast alone. Second, the mean of the analysis is a weighted average between the data, $Y_t$ and the forecast $mu_{f,t}$, each weighted by their precision. This averaging can also be viewed as a 'nudge', pushing the forecast toward the new data, with the magnitude of that nudge determined by the uncertainties in both the forecast and data (i.e. highly precise data = big nudge, highly uncertain data = very small nudge).

@@

```{r enkf, echo=FALSE, fig.cap="Example assimilation of a disturbance event shows that conventional assimilation approaches, such as EnKF, take time to converge to new states after disturbance (orange). By modeling the probability of disturbance (black) our Multinomial assimilation (blue) can correctly jump between states during disturbance events (e.g. P(disturbance) = 0.977 at time 10) while remaining resilient to potential false-positive events (e.g. time 12 and 15)."}
SSSD_figure <- function(){
  load("SSSD.RData")      ## simulation from SingleSite_SingleDisturbance.Rmd
  plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
     ylab="LAI",xlab="time")
  polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
    an0.stats[1, 1])), border = NA, col = '#FFBF6C')   ## EnKF interval
  ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256)) ## Multinomial interval
  lines(an0.stats[3,],col="orange",lwd=2)   ## EnKF median
  lines(CI.fN[3,],col="blue",lwd=2)         ## Multinomial median
  lines(Dbar,lwd=2)                         ## Multinomial disturbance rate
  points(data$Y,pch=18,col=2)               ## data
  legend("topright",legend=c("EnKF","Multinomial","data","P(dist)"),
       lwd=5,col=c("orange","blue","red","black")) ##"#b2df8a"
}
SSSD_figure()
```

** TODO from zoey for above figure: Labels could be more informative (i.e. EnKF = conventional and Multinomial = improved approach). Icon for "data" should match the icons rather than a line. Is the P(dist) bound at 0,1? I barely noticed the black line at first, it could maybe go in a second panel below with a shared x-axis and a distinct y-axis.**

** TODO from zoey for below figure: I think this figure could be more streamlined to better emphasize the difference between the filters. Why change the assimilation color from orange to blue in this figure?   The histogram, to me, seems more intuitive for displaying the data values rather than the forecasted values. The second panel doesn't have "forecast" in the color labels which also adds to confusion. The X axis could communicate the low-LAI = disturbed, high-LAI = undisturbed, without relying on the colors/legend. X axis for this and the Y axis for the above figure could just be "disturbance magnitude" **

```{r concept, echo=FALSE, message = FALSE,error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF, a multimodal forecast (gray histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabilities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

## EnKF
hist(LAI, probability = TRUE, main = "Conventional EnKF")#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))

@@

hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))), main="Multinomial Filter")
lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(density(as.matrix(jdat)),col="blue",lwd=3)
legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))
```

@@

To address this fundamental flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges. To begin, let's use $X$ to represent the latent (unobserved) "true" state of the system. Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent ($L$) continuous state variables of interest, which we'll refer to as $X^L_{k,t}$. Similar to the EnKF we'll assume that the ensemble forecast for any one of these categories can be represented as a multivariate Normal,

@@

where $\vec{\mu}_{f,k,t}$ and $P_{f,k,t}$ are the ensemble forecast ($f$) sample mean and covariance for outcome $k$ at time $t$. However, because we have $k$ categories we end up with a multimodal prediction (i.e., a mixture of MVNs). Furthermore, these modes need not be distinct; it is perfectly fine in concept for these MVNs to be overlapping. In practice the more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabilistically. Indeed, to be able to represent our multimodal prediction as a single probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.

@@

This is a generalization of the Bernoulli distribution (Binomial with n=1) and for any individual draw from the Multinomial $\vec{\rho_t}$ will be a vector of length $k$ with a 1 in the category being assigned and a 0 for all other categories.

@@

$$X_t = X^L_t \vec{\rho_t}$$
which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$. In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$. Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is no simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$. In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution. Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].

** TODO: Adjust fig 2 to make colors consistent between panels and with other figures**

@@

When making ensemble forecasts using the Analysis from the traditional EnKF, one needs to sample new initial conditions from the Analysis joint posterior distribution. This work fine if the initial conditions are complete (100% of the model variables represented) and when there is no covariance with any of the other uncertainties being propagated (e.g. sampling ensemble drivers or posterior parameter distributions). In practice, this assumption is frequently violated, for example an above average ensemble member might be above average not just because of its initial condition, but also because it had favorable parameters or drivers, creating a complex covariance between states, parameters, and drivers that would be destroyed by randomly resampling initial conditions. Because of this Anderson et al. [-@Anderson2001] developed an alternative know as _ensemble adjustment_ that nudges individual ensemble members toward the posterior analysis mean and covariance. Mathematically, the approach is essentially the multivariate generalization of using the prior forecast mean and variance to calculate a Z-score for every ensemble member (i.e. express as an standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.

In our Multinomial filter it is important to generalize this concept of ensemble adjustment to both discretely move ensemble members from one category to another, and then to preserve the covariances among state variables, and with drivers and parameters, when updating the discrete categorical assignments of individual ensemble members. To reassign forecasted ensemble members to new disturbance classes we first identify the "sink" categories that increased in probability between the prior, $p_{k,t}$, and the posterior, $\bar{\rho}_{k,t}$.

@@

$$c_{A|k} = Multinomial\left( n_k, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$

** TODO: Check equation formatting **

Conditional on this reassignment, the continuous states are updated according to [@Anderson2001] using the posterior statistics for that $X^L_{k,t}$. It is also worth noting that some cases of class reassignment may also require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well. One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT, with each ensemble member constructed with the identical anomalies around each PFT's mean..

@@

For the second experiment, we used the VSEM to extend the Multinomial assimilation model to a single site, single disturbance example with multiple pools. In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range. The goal here was not to replicate any specific system, but to provide an amount of parameter uncertainty that was realistic but not so large as to generate trivial ensemble forecasts. Model runs were for 25 years using VSEM's default simulated PAR as the driver and with data assimilation occurring annually. Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment. Process error was added to the simulation annually assuming a zero-bound tobit distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools. Ensemble members experienced disturbance stochastically with a 5% annual probability. For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 $kgC/m^2$ for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).

@@

Our fourth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT. Specifically PFT 2 was identical to the VSEM default, but PFT 1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT. In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 to PFT2, while disturbance 2 (defoliation) keeps the current PFT is. As with the previous scenario, the prescribed disturbance in the pseudodata was type 1 (stand clearing, transition from forest to shrub).

Our fifth and final experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other. To keep things simpler, this experiment reverted to a single disturbance type, same as experiment 2, except we dropped the disturbance probability to 2.5%. This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [@Dokoohaki2022].

@@

** TODO: TRY AGAIN AT GETTING MAPS ON SAME PROJECTION **

Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades. In summary, we: (1) calibrated VSEM against field data, then (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transition, then (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types. We then assessed the ability of the algorithm to detect different disturbances types and magnitudes.

### Study region and data sources

Our study region was a portion of central Oregon, USA that includes the Ameriflux tower used later to initialize the VSEM (see next section).  We used biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.  Available on a tile-basis (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html), we downloaded these data for tile h03v04, which corresponds to the region between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. All three products were 30m resolution and annual timestep;biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to be the same as biomass and cover. The biomass product was constructed by linking Landtrendr-processed Landsat data with field plot data using the gradient nearest neighbor (GNN) methodology [@Ohmann2002] to produce annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To simplify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). 
The disturbance product linked LandTrendr-derived disturbance information [@KENNEDY2010] with photo-interpreted labels of disturbance to attribute drivers to each disturbance event [@KENNEDY2015].  
To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances"}
## maps of pre-disturbance biomass, cover, and 2004 disturbance

@@

## Cover
if(!exists("cov04")) cov04 = raster::raster("cov04.tif")
load("rat.RData")
raster::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover",
            axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4))
#cov04 = terra::rast("cov04.tif")
#terra::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover")

>>>>

@@

The Ensemble Kalman Filter (EnKF) [@Evensen2009; @Evensen2009a] is a popular data assimilation algorithm that we use to demonstrate the challenges conventional data assimilation approaches have with capturing discrete disturbances and as our jumping off point for developing more flexible methods. Like most data assimilation algorithms the EnKF involves two cyclic steps: a Forecast step where a model is used to probabilistically predict the system state (and any other outputs) forward in time, and an Analysis step where new observations are used to update system state variables (e.g. carbon pools) and sometimes model parameters (Figure \@ref(fig:conceptualFig)). These updated variables are then used as the initial conditions for the next Forecast round.

Within the EnKF, the probabilistic Forecast step is performed using an ensemble of model runs that vary in at least initial conditions, but potentially also model parameters, drivers, and process error. The ensemble forecast approach is easy to apply to any model, is particularly good at capturing nonlinear predictions, and unlike many other approaches does not require solving for the adjoint of the model. In EnKF the multivariate forecast, $f$, at any point in time, $t$, is then summarized in terms of a vector of means across ensemble members, $\mu_{f,t}$ and the sample covariance, $P_{f,t}$. In our example we are interested in the three state variables in the VSEM, ${C_l,C_w,C_s}$, at a single site and assume an ensemble size of 200. Therefore $\mu_{f,t}$ is a vector of length 3 and $P_{f,t}$ is a $3 \times 3$ covariance matrix. If we were making a spatial forecast at $n_s$ locations, $\mu_{f,t}$ would be a vector of length $3 n_s$ and $P_{f,t}$ would be a $3 n_s \times 3 n_s$ covariance matrix that would capture the full set of correlations across both variables and space. An important difference between data assimilation and spatial statistics is that in data assimilation the spatial covariance matrix comes from (1) the mechanistic understanding embedded in our process model and (2) the spatial covariance in the forecast inputs (initial conditions, parameters, drivers, process error), rather than an empirical semivariogram.

The Analysis step in the EnKF is the same as the standard Kalman Filter. Because the forecast step from t-1 to t occurs prior to the observed data at time t, $Y_t$, the forecast of our three state variables is used as an informative _prior_ distribution and is assumed to follow a multivariate Normal (MVN) distribution with mean $\mu_{f,t}$ and covariance $P_{f,t}$. The EnKF also acknowledges that the observed data at time t, $Y_t$, is measured imperfectly. Observation error is assumed to follow a MVN distribution with an observation error covariance matrix $R$. Furthermore, it is assumed that $R$ is known, which is not unreasonable as many derived data products and sensors report uncertainties, and uncertainties from field data can often be inferred from sampling theory or Monte Carlo resampling. In many cases new data do not exist for every state variable (e.g. we may observe leaves and wood but not soil C) and so an observation operator matrix, $H$ is also introduced, which in most cases is just a $n \times m$ indicator matrix matching the observed $Y$'s (n=2 in this example) with the vector of state variables (m=3 in this example). Indeed, one of the powerful features of data assimilation is its ability to update unobserved state variables based on the covariances between observed and unobserved variables. Putting this all together, Bayes' Theorem can then be used to update our MVN forecast prior, given a MVN Likelihood of $Y_t$ with covariance $R$, giving an analytically-tractable conjugate MVN posterior with mean $\mu_{a,t}$ and covariance $P_{a,t}$:

@@

The subscript $a$ is used to refer to the Analysis posterior and distinguish this from the prior forecast $f$ mean and covariance for the same timestep. While the matrix notation can make these equations look complicated, the underlying meaning is accessible. First, the precision of the analysis (i.e. 1/variance) is the sum of the observation precision and the forecast precision, meaning that the analysis is always more precise than the data or forecast alone. Second, the mean of the analysis is a weighted average between the data, $Y_t$ and the forecast $mu_{f,t}$, each weighted by their precision. This averaging can also be viewed as a 'nudge', pushing the forecast toward the new data, with the magnitude of that nudge determined by the uncertainties in both the forecast and data (i.e. more precise data = larger nudge).

@@

```{r enkf, echo=FALSE, fig.cap="Example assimilation of a disturbance event shows that conventional assimilation approaches, such as EnKF, take time to converge to new states after disturbance (orange). By modeling the probability of disturbance (black) our Multinomial assimilation (blue) can correctly jump between states during disturbance events (e.g. P(disturbance) = 0.977 at time 10) while remaining resilient to potential false-positive events (e.g. time 12 and 15)."}
SSSD_figure <- function(){
  load("SSSD.RData")      ## simulation from SingleSite_SingleDisturbance.Rmd
#  layout(matrix(c(1,1,1,2),nrow=4,ncol=1))
  plot(data$Y,ylim = range(rbind(CI.fN,data$Y)),xlim=xlim,type='n',
     ylab="LAI",xlab="time")
  polygon(cbind(c(time, rev(time), 1), c(an0.stats[2, ], rev(an0.stats[4, ]),
    an0.stats[1, 1])), border = NA, col = '#FFBF6C')   ## EnKF interval
  ecoforecastR::ciEnvelope(1:NT,CI.fN[2,],CI.fN[4,],col=rgb(col[1],col[2],col[3],0.3*256,maxColorValue=256)) ## Multinomial interval
  lines(an0.stats[3,],col="orange",lwd=2)   ## EnKF median
  lines(CI.fN[3,],col="blue",lwd=2)         ## Multinomial median
  lines(Dbar,lwd=3)                         ## Multinomial disturbance rate
  points(data$Y,pch=18,col="red")               ## data
  legend("topright",legend=c("EnKF","Multinomial","data","P(dist)"),
       lwd=c(5,5,NA,5),pch=c(NA,NA,18,NA),col=c("orange","blue","red","black")) ##"#b2df8a"
#  plot(Dbar,lwd=2,ylab="P(dist)")
}
SSSD_figure()
```

```{r concept, echo=FALSE, error=FALSE, fig.asp=1, fig.cap="Top: In a conventional EnKF  a multimodal forecast (gray histogram) is approximated by a unimodal Normal prior (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the Multinomial filter the same multimodal forecast is represented by a mixture of Normal priors (green) each weighted by their prior (forecast) probabilities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred.", message=FALSE, warning=FALSE,results='hide'}
library(rjags)
par(mfrow=c(2,1))
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

@@

## EnKF
hist(LAI, probability = TRUE, main = "Conventional EnKF")#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
#legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","blue"))

@@

hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))), main="Multinomial Filter")
lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="#b2df8a",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(density(as.matrix(jdat)),col="blue",lwd=3)
legend("top",legend=c("ensemble","prior","data","EnKF","Multinomial"),
       lwd=3,col=c("gray","#b2df8a","red","orange","blue"))
```

@@

To address this flaw in the Kalman Analysis, we need to build into the Analysis step both the possibility for ensemble forecasts to be multimodal, and the ability for the data assimilation to jump discretely between modes all-at-once rather than through gradual nudges. To begin, let's use $X_t$ to represent the latent (unobserved) "true" state of the system at time $t$. Furthermore, we will consider a process-model that is capable of predicting $k$ different categorical outcomes (land cover classes, types of disturbance, etc.) each of which has its own prediction for the latent ($L$) continuous state variables of interest, which we'll refer to as $X^L_{k,t}$. Similar to the EnKF we'll assume that the ensemble forecast for any one of these categories can be represented as a multivariate Normal,

@@

where $\vec{\mu}_{f,k,t}$ and $P_{f,k,t}$ are the ensemble forecast ($f$) sample mean and covariance for outcome $k$ at time $t$. However, because we have $k$ categories we end up with a multimodal prediction (i.e., a mixture of MVNs). Furthermore, these modes need not be distinct; in concept it is perfectly fine for these MVNs to be overlapping. In practice the more they overlap the harder it is to identify which mode you are in, but as we will see this categorical assignment will be treated probabilistically. Indeed, to be able to represent our prediction as a single multimodal probability distribution we need to know the _forecasted_ probability of being in each category, which we'll take as known and represent using the vector $\vec{p_t}$, giving us a weighted MVN mixture as our forecast.

@@

This is a generalization of the Bernoulli distribution (Binomial with n=1). For any individual draw from the Multinomial, $\vec{\rho_t}$ will be a vector of length $k$ with a 1 in the category being assigned and a 0 for all other categories.

@@

$$X_t = X^L_t \vec{\rho_t}$$

which for any individual draw of $\vec{\rho_t}$ assigns $X_t$ that specific $X^L_{k,t}$. In practice, we will use Bayes' Theorem to integrate over the uncertainties in both our categorical assignment and continuous state variables, producing posterior Analysis probability distributions for both $X_t$ and $\vec{\rho_t}$. Unlike in the EnKF, where conjugacy allows us to calculate an analytical solution for the posterior, there is no simple analytical solution for the joint posterior distribution of $X_t$ and $\vec{\rho_t}$. In this case we will use well-developed Bayesian Markov Chain Monte Carlo (MCMC) numerical algorithms to sample from the posterior distribution. Specifically, numerical analyses in this paper were conducted using JAGS v4.2.0 [@Plummer03] accessed from R v4.0.3 [@Rcore] using the rjags package v4.10 [@rjags].

@@

When using the EnKF Analysis to initialize the next forecast, one needs to sample a new ensemble of initial conditions from the Analysis joint posterior distribution. This works fine if there is no covariance with any of the other uncertainties being propagated (e.g. sampling ensemble drivers or posterior parameter distributions) and the initial conditions are complete (100% of the model variables represented). In practice, this assumption is frequently violated, for example an above average ensemble member might be above average not just because of its initial condition, but also because it had favorable parameters or drivers, creating a complex covariance between states, parameters, and drivers that would be destroyed by randomly resampling initial conditions. Because of this Anderson et al. [-@Anderson2001] developed an alternative know as _ensemble adjustment_ that nudges individual ensemble members toward the posterior analysis mean and covariance. Mathematically, the approach is essentially the multivariate generalization of using the prior forecast mean and variance to calculate a Z-score for every ensemble member (i.e. a standard Normal anomaly with mean 0 and standard deviation 1), and then apply that Z-score to the new posterior analysis mean and variance.

In our Multinomial filter it is important to generalize this concept of ensemble adjustment to both discretely move ensemble members from one category to another, and then to preserve the covariances among state variables and with drivers and parameters, when updating the discrete categorical assignments of individual ensemble members. To reassign ensemble members to new disturbance classes we first identify the "sink" categories that increased in probability between the prior, $p_{k,t}$, and the posterior, $\bar{\rho}_{k,t}$.

@@

$$c_{A|k} = Multinomial\left( n_k,\, \frac{\vec{\delta p_t}}{\sum \vec{\delta p_t}}  \right)$$

Conditional on this reassignment, the continuous states are updated according to [@Anderson2001] using the posterior statistics for that $X^L_{k,t}$. It is also worth noting that some cases of class reassignment may require keeping track of categorical switches in Plant Functional Type (PFT), such as a transition from forest to agriculture, which implies a large jump in PFT parameters as well. One could apply the same [@Anderson2001] style adjustment to PFT parameters, but for simplicity here we just maintained alternative parameter ensembles for each PFT, with each ensemble member constructed with the identical anomalies around each PFT's mean.

@@

For the second experiment, we used the VSEM to extend the single site, single disturbance example to a model with multiple pools. In this and the following simulated data experiments we used VSEM's default parameters, drawing ensemble members from Normal distributions centered on the prior mean and assuming a standard deviation that was 1/10th the default uniform prior range. The goal here was not to replicate any specific system, but to provide an amount of parameter uncertainty that was realistic but not so large as to generate trivial ensemble forecasts. Model runs were for 25 years using VSEM's default simulated PAR as the driver and with data assimilation occurring annually. Model initial conditions were generated by an initial 25 year ensemble spin-up prior to the 25 year experiment. Process error was added to the simulation annually assuming a zero-censored Normal (a.k.a. tobit) distribution with a process error covariance matrix that was diagonal with variances of 0.1 for all pools. Ensemble members experienced disturbance stochastically with a 5% annual probability. For the disturbance itself, the post-disturbance leaf and wood carbon were drawn from a tobit with a mean of 1 $kgC/m^2$ for both pools, standard deviations of 0.15 and 0.25 respectively, and 25% of the biomass lost was assumed to enter the soil C pool, with the remainder removed (which could be thought of as either atmospheric emissions or biomass harvest depending on the disturbance type).

@@

Our fourth experiment was similar to the third, but extended the single-site, multi-disturbance, multi-pool scenario to one where one disturbance type included a switch in PFT. Specifically PFT2 was identical to the VSEM default, but PFT1 was parameterized with a stem residence time that was increased by 10x to create a more tree-like PFT. In addition, we established a transition matrix where disturbance 1 (stand clearing) transitions from PFT1 (forest) to PFT2 (shrub), while disturbance 2 (defoliation) keeps the current PFT. The prescribed disturbance in the pseudodata was type 1 (stand clearing).

Our fifth experiment extended the analysis to a two-site scenario with spatial covariance across sites and pools, but with a disturbance occurring in one site but not at the other. To keep things simpler, this experiment reverted to a single disturbance type, same as experiment 2, except we dropped the disturbance probability to 2.5%. This scenario was designed to test a concern we had in previous research that a disturbance occurring at one site could cause the biomass to decline at other undisturbed sites because of the cross-site covariances [@Dokoohaki2022].

@@

Having tested the Multinomial assimilation against simulated data we next applied the algorithm to multiple remotely-sensed disturbance types for an evergreen forested region in the central Oregon Cascades, USA. In summary, we: (1) calibrated VSEM against field data; (2) used a remotely-sensed disturbance product to construct priors for disturbance rates and PFT transitions; (3) extracted and assimilated ~400 time-series of remotely sensed aboveground biomass, stratified by different disturbance types; and then (4) assessed the ability of the algorithm to detect different disturbances types and magnitudes. Model calibration is discussed in the Supplementary material.

### Remotely-sensed data sources

Data for our study region include biomass, cover, and disturbance data derived from time-series Landsat imagery processed with the LandTrendr algorithm under a separate project.  Available on a tile-basis (http://emapr.ceoas.oregonstate.edu/pages/data/viz/index.html), we downloaded these data for tile h03v04, which corresponds to the region between 120 47'W and 123 7'W longitude and 43 45'N and 45 '4'N latitude. All three products were 30m resolution and annual timestep; biomass and cover extended from 1990-2017, while disturbance extended from 1984-2011 and a nearest neighbor algorithm was used to regrid the disturbance data to the same projection as biomass and cover. The biomass product was constructed by linking Landtrendr-processed Landsat data with field plot data using the gradient nearest neighbor (GNN) methodology [@Ohmann2002] to produce annual biomass estimates [@Kennedy2018]. The cover product trained a spatiotemporal exploratory model (STEM), which uses an ensemble of regression trees defined within spatially overlapping support sets, against the 2001 U.S. National Land Cover Database (NLCD) to produce annual land-cover maps for Washington, Oregon, and California [@HOOPER2018]. To simplify the analysis we focused on a single vegetation type, evergreen forest, which is the dominant vegetation cover in the region and consists primarily of Ponderosa pine (_Pinus ponderosa_). 
The disturbance product linked LandTrendr-derived disturbance information [@KENNEDY2010] with photo-interpreted labels of disturbance to attribute drivers to each disturbance event [@KENNEDY2015].
To simplify analyses we aggregated the disturbance product into five classes: undisturbed, cut, fire, pest, and other.

```{r map, echo=FALSE, fig.asp=1, fig.cap="Remotely sensed observations from central Oregon for 2004 used as inputs to test the multinomial filter against real-world disturbances. Plots are in Albers Equal Area projection with axes in meters and cover a bounding box spanning roughly 43.5 to 45 North latitude by -121 to -123 West longitute and"}
## maps of pre-disturbance biomass, cover, and 2004 disturbance

@@

## Cover
if(!exists("cov04")) cov04 = raster::raster("cov04.tif")
if(!exists("dis04"))  dis04 = raster::raster("dis04.tif")
load("rat.RData")
if(!exists("cov04b")){
  #cov04b = raster::projectRaster(cov04,crs=crs(dis04))
  #save(cov04b,file="cov04b.Rdata")
  load("cov04b.Rdata")
}
raster::plot(cov04b,col=rat$color,asp=1,main="NLCD Land Cover",
            axis.args=list(at=seq_along(rat$description),labels=rat$description,cex.axis=0.4))
#cov04 = terra::rast("cov04.tif")
#terra::plot(cov04,col=rat$color,asp=1,main="NLCD Land Cover")
